Omics-based Single Step Trait Prediction

### 2017-09-06 paper/manuscript.tex
paper, update results

\section*{Results}
\Genetics2level{Inbred lines}
At the level of the 149 genotypes for which both, genomic (G) and 
transcriptomic (T) information, were available, the use of T yielded higher
predictive abilities than the use of G for four out of six traits
(Fig.~\ref{fig:InbredResults}A).
In terms of predictive ability, the absolute advantage of T over G was as large 
as 0.04 as observed for the traits DS, ED, GW and KW\@.
Except for KW, the predictive ability for predictions based on G were always
higher for the full set of inbred lines ($n = 211$) than for the reduced set of 
inbred lines ($n = 149$).
The combination of G and T (\textit{i.e.}, GT) in a single-step approach for the 
prediction of all 211 lines, performed at least as well as G alone when T was 
superior over G for the reduced set of 149 lines.
In the case of GW and ED, the use of GT yielded a slight improvement in predictive 
ability over G.
Where T was a worse predictor for the reduced set of 149 lines than G,
predictive abilities of GT for the full set of 211 lines were also worse than 
those based on G alone.

@@

\Genetics2level{Hybrids}
The maize hybrid data set contains information on six agronomic traits and
three predictors, namely pedigree data (P), genomic information (G), 
transcriptomic information (T), and combinations thereof.
Pedigree records and genomic information were available for the 245 parent 
lines of all 1,521 hybrids and we denote these data as the `Full' set whereas
transcriptomic data were available for 103 parent lines of 685 hybrids, which
we denote as `Core'.
For DMY, T was the best predictor in the core set of 685 hybrids with a 
predictive ability that was 0.08 points higher than for genomic data, as the 
second best predictor for this trait (Fig.~\ref{fig:HybridResults}).
In the core set, P had the highest predictive ability for STA and DMC\@.
For the other three traits, G was the superior predictor inside the core set.
In the full set of 1,521 hybrids, predictive abilities obtained with G rose 
markedly for all traits with the exception of FAT\@.
The combination of genomic and transriptomic information (GT) in a single-step
model yielded a small improvement over G alone for the traits DMY and PRO\@.
Except for PRO, compared to using only G, trait prediction did not benefit from 
the single step approach when T was not superior to G in the core set.
Predictive abilities resulting from combinations of pedigree with genomic (PG)
information were better than P alone for all traits and PT improved upon P in
the core set for four traits.

@@

\Genetics2level{Coverage of the genetic space}
Single-step approaches are based on the use of one complete predictor that
covers all genotypes and one incomplete predictor that covers only a subset of
all available genotypes.
Here, we explore the impact of which data are covered by the incomplete 
predictor on the predictive abilities obtained via single-step approaches by
focusing exclusively on the 149 inbred lines from the maize diversity panel
that are covered by both, genomic and transcriptomic information.
First, we assembled nine different core sets, which we defined as subsets
of genotypes that maximize the average genetic distance between genotypes of
the full set.
The nine core sets differ in that each covers a different number of genotypes,
ranging from only 10\% to 90\% of all genotypes covered by the incomplete
predictor (Fig.~\ref{fig:inbred-pca}).

@@

Predictive abilities obtained with the different core sets varied widely within
traits (Fig.~\ref{fig:InbredResults}B).
For the traits CW and PH, which were better predicted via G than via T in the 
reduced set of lines, predictive abilities were higher when only a small number 
of genotypes was included in the core set.
This corresponds to a situation where a large number of genotypes is being 
imputed and the majority of the information is derived from genomic information. 
For most other traits, an initial increase in predictive abilities when moving
from a core set size of 0.1 to 0.5/0.6 was observed, followed by a drop around
0.6/0.7 and a subsequent hike up to a core set size of 0.9.
A principal component analysis depicts the coverage of the genetic space by
genotypes (Fig.~\ref{fig:PopStructure}).
With a core set size of 50--60\% of the 149 genotypes, the congruence between
core set and complement was high whereas genotypes in the south-west direction
of the PCA plot were overrepresented compared to genotypes in the east 
direction at a core set size of 0.7.

As an alternative to exploring the impact of genotypes sampled based on the
maximization of genetic distance within the core set, we wanted to explore a
more extreme scenario where we select core sets based on population structure.
Hereto, we ran an admixture analysis on the same set of 149 inbred lines 
assuming $K=3$ ancestral populations (Fig.~\ref{fig:PopStructure}).

Of the 149 genotypes, 17 belonged to more than 50\% to the first ancestral
population (A1), 109 belonged to more than 50\% to the second ancestral
population (A2) and 19 belonged to more than 50\% to the third ancestral
population while four genotypes could not be unequivocally assigned to either
ancestral population (mixed).
To reconcile the large size of the A2 set compared to A1 and A3, we selected 19 
genotypes from A2 using previously described core sampling.
For single-step prediction, each of the three genotype sets were then declared
as having genomic and transcriptomic information whereas their complements were
covered by transcriptomic data only.
Genotypes from A1 and A3, which were at the far ends in the PCA plot 
(Fig.~\ref{fig:PopStructure}B), were estimated with less precision than genotypes
from the central cluster of genotypes belonging largely to the A2 population
(Fig.~\ref{fig:PopStructure}A, Fig.~\ref{fig:InbredResults}C).

Regarding the hybrid material, a PCA carried out separately for the two groups
of parental inbred lines showed that the entire genetic space in each group 
was covered well by genotypes that had both, transcriptomic and genomic data 
(Fig.~\ref{fig:PCA}).

>>>>

\section*{Results}
\Genetics2level{Inbred lines}
At the level of the 110 genotypes for which both, genomic (G) and 
transcriptomic (T) information, were available, the use of T yielded higher
predictive abilities than the use of G for two out of six traits
(Fig.~\ref{fig:InbredResults}A).
In terms of predictive ability, the absolute advantage of T over G was as large 
as 0.15 as observed for the trait ED\@.
The predictive ability for predictions based on G were always higher for the
full set of inbred lines ($n = 164$) than for the reduced set of inbred lines
($n = 110$).
The combination of G and T (\textit{i.e.}, GT) in a single-step approach for the 
prediction of all 164 lines, performed at least as well as G alone when T was 
superior over G for the reduced set of 110 lines.
In the case of KW and ED, the use of GT yielded a slight improvement in predictive 
ability over G.
Where T was a worse predictor for the reduced set of 110 lines than G,
predictive abilities of GT for the full set of 164 lines were also worse than 
those based on G alone.

@@

\Genetics2level{Hybrids}
The maize hybrid data set contains information on seven agronomic traits and
three predictors, namely pedigree data (P), genomic information (G), 
transcriptomic information (T), and combinations thereof.
Pedigree records and genomic information were available for the 245 parent 
lines of all 1,521 hybrids and we denote these data as the `Full' set whereas
transcriptomic data were available for 103 parent lines of 685 hybrids, which
we denote as `Core'.
In the `core' set, transcriptomic data were the best single predictor for ADL,
PRO and particulaly for DMY where the predictive ability obtained with T
was 0.08 points higher than for G as the second best predictor for this trait
(Fig.~\ref{fig:HybridResults}).
In the core set, P had the highest predictive ability for STA and DMC\@.
For the FAT and SUG, G was the superior predictor inside the core set.
In the full set of 1,521 hybrids, predictive abilities obtained with G rose 
markedly for all traits with the exception of FAT\@.
The combination of genomic and transriptomic information (GT) in a single-step
model yielded a small improvement over G alone for the traits DMY and PRO\@.
Except for PRO, compared to using only G, trait prediction did not benefit from 
the single step approach when T was not superior to G in the core set.
Predictive abilities resulting from combinations of pedigree with genomic (PG)
information were better than P alone for all traits and PT improved upon P in
the `full' set for four traits.

@@

\Genetics2level{Impact of coverage by the incomplete predictor}
Single-step approaches are based on the use of one complete predictor that
covers all genotypes and one incomplete predictor that covers only a subset of
all available genotypes.
Here, we explore the impact of which data are covered by the incomplete 
predictor on the predictive abilities obtained via single-step approaches by
focusing exclusively on the 110 inbred lines from the maize diversity panel
that are covered by both, genomic and transcriptomic information.
First, we assembled nine different core sets, each with 100 nested subsamples,
which were generated by randomly removing transcriptomic information from
some genotypes (Fig.~\ref{fig:compound-plot}).
The nine core sets differ in that each covers a different number of genotypes,
ranging from only 10\% to 90\% of all genotypes covered by the incomplete
predictor (Fig.~\ref{fig:inbred-pca}).

@@

Predictive abilities obtained with the different core sets varied widely within
most traits (Fig.~\ref{fig:InbredResults}).
For most traits, an initial increase in predictive abilities when moving
from a core set size of 0.1 to 0.2/0.3 was observed, regardless of whether G
or T individually was the superior predictor.
For the traits CW, GW and PH, which were better predicted via G than via T in
the reduced set of lines, predictive abilities were higher when only a small
number of genotypes was included in the core set.
This corresponds to a situation where a large number of genotypes is being 
imputed and the majority of the information is derived from genomic information.
A hike in predictive abilities, when moving from a core fraction of 10\% to 
90\%, was observed for ED and KW, which were predicted better when using T
individually compared to using G individually.
The only trait that was barely affected by a change in the core fraction was DS
for which the predictive abilities using either G or T individually were almost
identical (Fig.~\ref{fig:InbredResults}).
